The zoon R package for reproducible and shareable species distribution modelling

### 2016-12-24 ms.Rmd
sorted out margins in maxnet plot

```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
mosquito1 <- workflow(occurrence = UKAnophelesPlumbeus,
                      covariate  = UKBioclim,
                      process    = Background(n = 500),
                      model      = MaxEnt,
                      output     = InteractiveMap)
```

@@

@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the zoon data into the spThin format, uses this package to apply the algorithm, and converts data back into zoon's expected format again:

@@

zoon enables us to investigate this hypothesis by easily comparing MaxEnt models fitted with and without regularisation and feature construction.
We fit MaxEnt with both the usual java implementation as well as the downweighted logistic regression approximation provided in the `maxnet` R package [@maxnet].
We also compare these models with a standard logistic regression model which does not apply the downweighting step.
The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@Royle2012; @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.

```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE, fig.height = 7, fig.width = 9}
maxnet_comparison <- workflow(
  occurrence = CarolinaWrenPO,
  covariate  = CarolinaWrenRasters,
  process    = Chain(SubsampleOccurrence(500),
                     Background(n = 10000),
                     CarolinaWrenValidation),
  model      = list(MaxEnt(args = 'threshold=false'),
                    MaxNet,
                    MaxNet(regmult = 0.005),
                    MaxNet(features = 'l'),
                    MaxNet(regmult = 0.005, features = 'l'),
                    LogisticRegression),
  output     = Chain(PrintMap(points = FALSE),
                     PerformanceMeasures))
```

```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 6, fig.width = 9}

# plot the figure to disk, then plot in a separate chunk with a legend
par(mfrow = c(3, 2),
    mar = c(6, 5, 8, 5))

# loop through plotting them
for (i in 1:6) {
  fpath <- sprintf('figs/maxnet_comparison-%i.png', i)
  img <- brick(fpath)
  plotRGB(img)
  mtext(LETTERS[i],
        side = 3,
        line = -1.5,
        adj = 0)
}

@@

During the development of zoon we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
In order to facilitate a user –friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.

>>>>

```{r plotworkflow, eval = FALSE, echo = TRUE}
mosquito1 <- workflow(occurrence = UKAnophelesPlumbeus,
                      covariate  = UKBioclim,
                      process    = Background(n = 500),
                      model      = MaxEnt,
                      output     = InteractiveMap)
```

@@

@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data by `thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the zoon data into the spThin format, uses this package to apply the algorithm, and converts data back into zoon's expected format again:

@@

zoon enables us to investigate this hypothesis by easily comparing MaxEnt models fitted with and without regularisation and feature construction.
We fit MaxEnt with both the usual java implementation as well as the downweighted logistic regression approximation provided in the `maxnet` R package [@maxnet].
We also compare these models with a standard logistic regression model which does not apply the downweighting step.
The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@Royle2012; @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
Note that setting the regularisation constant in the `maxnet` package to 0 caused numerical errors, so we use a very small value instead.

```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE, out.height = 600, out.width = 600}
maxnet_comparison <- workflow(
  occurrence = CarolinaWrenPO,
  covariate  = CarolinaWrenRasters,
  process    = Chain(SubsampleOccurrence(500),
                     Background(n = 10000),
                     CarolinaWrenValidation),
  model      = list(MaxEnt(args = 'threshold=false'),
                    MaxNet,
                    MaxNet(regmult = 0.005),
                    MaxNet(features = 'l'),
                    MaxNet(regmult = 0.005, features = 'l'),
                    LogisticRegression),
  output     = Chain(PrintMap(points = FALSE),
                     PerformanceMeasures))
```
```{r maxnet_buffer, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", out.height = 7, out.width = 7}

# loop through adding a margin to each of the previous figures
for (i in 1:6) {
  fpath_in <- sprintf('figs/maxnet_comparison-%i.png', i)
  fpath_out <- sprintf('figs/maxnet_buffer-%i.png', i)
  img <- brick(fpath_in)
  ext <- as.vector(extent(img))

  png(fpath_out,
      width = ncol(img),
      height = nrow(img))

  par(oma = rep(0, 4), mar = rep(0, 4))
  plot.new()
  plot.window(xlim = ext[1:2],
            ylim = ext[3:4],
            asp = 1)
  plotRGB(img, maxpixels = Inf, add = TRUE)

  dev.off()
}
```

```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 7, fig.width = 9}

# plot the figure to disk, then plot in a separate chunk with a legend
par(mfrow = c(3, 2))

# loop through plotting them
for (i in 1:6) {
  fpath <- sprintf('figs/maxnet_buffer-%i.png', i)
  img <- brick(fpath)
  plotRGB(img, maxpixels = Inf)
  mtext(LETTERS[i],
        side = 3,
        line = -1.5,
        adj = 0)
}

@@

During the development of zoon we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
In order to facilitate a user–friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.
